---
title: "WorkFlow Basics"
author: "Gabriel Montes"
date: "2025-07-17"
output: pdf_document
---

### 1. Why does this code not work?

```{r}
my_variable <- 10
my_variable
#> Error: object 'my_varıable' not found
```

make sure the variable names are the same

### 2. Tweak each of the following R commands so that they run correctly:

```{r}
library(tidyverse)
library(ggplot2)
ggplot(data = mpg) +
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy), method = "lm")

```

### 3. Press Option + Shift + K / Alt + Shift + K. What happens? How can you get to the same place using the menus?

opens the rstudio commands shortcuts

### 4. Let’s revisit an exercise from the Section 1.6. Run the following lines of code. Which of the two plots is saved as mpg-plot.png? Why?

```{r}
my_bar_plot <- ggplot(mpg, aes(x = class)) +
  geom_bar()
my_scatter_plot <- ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
ggsave(filename = "mpg-plot.png", plot = my_bar_plot)
```

the first graph was saved since we saved it to a variable and specified in ggsave which to save
