---
title: "Data tidying"
author: "Gabriel Montes"
date: "2025-07-17"
output: pdf_document
---


### 1. For each of the sample tables, describe what each observation and each column represents.

in table 1
each row represents a unique combination of country and year
each column is a variable, country, year, cases, population

in table 2
each row represents a single measurement 
each column also are features of the measurement, country, year, type count

in table3
each row represents a country and year pair with the rate of TB cases 
each colum are the features, country year rate

### 2. Sketch out the process you’d use to calculate the rate for table2 and table3. You will need to perform four operations:

for table 2
```{r}
library(tidyverse)

table2_wide <- table2 |> 
  pivot_wider(names_from = type, values_from = count)

table2_rate <- table2_wide |> 
  mutate(rate = (cases / population) * 10000)
```

for table 3
```{r}
table3_separated <- table3 |> 
  separate(rate, into = c("cases", "population"), sep = "/") |> 
  mutate(
    cases = as.numeric(cases),
    population = as.numeric(population)
  )

table3_rate <- table3_separated |> 
  mutate(rate = (cases / population) * 10000)
```

